Extreme heterogeneity of influenza virus transcription in single cells

### 2017-09-14 paper/paper.tex
revisions to Discussion -- still need another pass over this before sending back to Alistair.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Extreme heterogeneity of influenza virus transcriptional load in single cells}

@@

\section{Introduction}
Viruses can cause massive and rapid changes in a cell's transcriptome as they churn out viral mRNAs and hijack cellular machinery.
For instance, cells infected with influenza virus at high multiplicity of infection (MOI) express an \emph{average} of 50,000 to 100,000 viral mRNAs per cell, corresponding to 5 to 25\% of all cellular mRNA~\citep{Hatada:1989vz}.
In addition, infection can trigger innate-immune sensors that induce the expression of cellular anti-viral genes~\citep{Killip:2015dw,Iwasaki:2014dw,Crotta:2013ef}.
This anti-viral response is another prominent transcriptional signature of high-MOI influenza virus infection in bulk cells~\citep{Geiss:2002a}.

However, initiation of an actual influenza infection typically involves just a few virions infecting a few cells~\citep{varble2014influenza,mccrone2017evolutionary}. 
The dynamics of viral infection in these individual cells may not mirror bulk measurements made on many cells infected at high MOI.
Over 70 years ago, \citet{Delbruck:1945wo} showed that there was a $\sim$100-fold range in the number of progeny virions produced per cell by clonal bacteria infected with clonal bacteriophage.
Subsequent work has shown similar heterogeneity in the number of progeny virions per cell during infection with other viruses~\citep{Zhu:2009ea,Schulte:2014br,combe2015single,Akpinar:2016gt}, including influenza virus~\citep{Heldt:2015iz}.

In the case of influenza virus infection, targeted single-cell measurements of specific proteins or RNAs have shed light on some factors that contribute to cell-to-cell heterogeneity. 
The influenza virus genome consists of eight negative-sense RNA segments, and many infected cells fail to express one more of these RNAs~\citep{Heldt:2015iz,Dou:2017cp} or their encoded proteins~\citep{Brooke:2013kb}.
In addition, the activation of innate-immune responses is inherently stochastic~\citep{Shalek:2013ej,Shalek:2014ey,bhushal2017cell}, and only some influenza-infected cells express anti-viral interferon genes~\citep{PerezCidoncha:2014jr,Killip:2017ef}.
However, the extent of cell-to-cell variation in these and other host and viral factors remains unclear, as does the association among them in individual infected cells.

@@

Infected cells will express viral as well as cellular mRNAs -- however the cell barcodes and UMIs cannot distinguish whether a cell was initially infected by one or multiple viral particles.
We therefore engineered an influenza virus (strain A/WSN/1933) that additionally carried \emph{viral barcodes} consisting of synonymous mutations near the 3' end of each transcript (Figure~\ref{fig:workflow}A).
Critically, these synonymous mutations did not greatly impact viral growth kinetics (Figure~\ref{fig:workflow}B).
We infected A549 human lung carcinoma cells with an equal mix of the wild-type and synonymously barcoded viruses.
Cells infected by a single virion will exclusively express mRNAs from either wild-type or synonymously barcoded virus, whereas cells that are co-infected with multiple virions will often express mRNAs from both the wild-type and synonymously barcoded viruses (Figure~\ref{fig:workflow}C).
\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{figures/Workflow/workflow.pdf}}
\caption{\label{fig:workflow} Experimental design.
{\bf (A)}  
We engineered a virus that carried two synonymous mutations near the 3' end of each mRNA.
At top are the specific mutations for PB2.
At bottom are locations of the synonymous mutations relative to the typical distribution of read depth for our 3'-end sequencing.
{\bf (B)} 
The wild-type and synonymously barcoded viruses transcribe their genes with similar kinetics. 
Shown is the abundance of the viral hemagglutinin (HA) transcript relative to the cellular housekeeping gene L32 as assessed by qPCR in A549 cells infected at an MOI of 0.5 (as determined on MDCK-SIAT1 cells).
Error bars $\pm$ S.D., n=3.
{\bf (C)}  
For the single-cell mRNA sequencing, A549 cells were infected with an equal mixture of mutant and wild-type virus. 
Immediately prior to collection, cells were physically separated into droplets and cDNA libraries were generated containing the indicated barcodes. 
The libraries were deep sequenced, and the data processed to create a matrix that gives the number of molecules of each transcript observed in each cell.
Infected cells were further annotated by whether their viral mRNAs derived from wild-type virus, synonymously barcoded virus, or both.
}
\figdata{Sequences of wild-type and barcoded viruses are in \texttt{viralsequences.fasta}.\label{figdata:viralsequences}}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/validating_populations_D02.pdf}}
\caption{\label{fig:viruspopulations} The viral stocks in our experiments are relatively pure of defective particles. 
{\bf (A)}
Our viral stocks have a higher ratio of infectious particles to HA virion RNA compared to a high-defective stock propagated at high MOI.
HA viral RNA was quantified by qPCR on virions. 
Error bars $\pm$ S.D., n=6 (qPCR replicates). 
{\bf (B)} 
Our viral stocks have a higher ratio of infectious particles to particles capable of expressing the viral HA protein.
A549 cells were infected with all three viral stocks at an MOI of 0.1, and the percentage of cells expressing HA protein at 9 hours post-infection was quantified by antibody staining and flow cytometry.
{\bf (C)} 
Our viral stocks are less immunostimulatory than virus propagated at high MOI. 
Measurements of \textit{IFNB1} transcript by qPCR normalized to the housekeeping gene L32 in A549 cells at 10 hours post infection at an MOI of 0.5.
Error bars $\pm$ S.D., n=3.
Note that the MOIs were calculated by TCID50 on MDCK-SIAT1 cells, whereas the experiments in this figure involved infection of A549 cells.
}
\figsupp[Full flow cytometry data for panel B . \label{figsupp:HAstain}]{
Full flow cytometry date for  Figure~\ref{fig:viruspopulations}B. 
A549 cells were infected at an MOI of 0.1 as calculated by TCID50 on MDCK-SIAT1 cells. 
{\bf (A)} Uninfected gating control.
{\bf (B)} Cells infected with the wild-type virus stock used in our experiments.
{\bf (C)} Cells infected with synonymously barcoded virus stock used in our experiments.
{\bf (D)} Cells infected with a stock of wild-type virus propagated at a high MOI, and therefore enriched in defective particles.
}{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/HA_stain_supplement.png}}\end{figure}

@@

\subsection{Absence of viral genes partially explains cell-to-cell variability in viral transcript abundance.}
The influenza genome is segmented, and cells can fail to express a viral mRNA if the encoding gene segment is not packaged in the infecting virion or fails to initiate transcription after infection.
Indeed, several groups have reported that the majority of infected cells fail to express at least one viral gene~\citep{Brooke:2013kb, Heldt:2015iz,Dou:2017cp}. 
We wondered if the absence of specific viral genes might be associated with reduced amounts of viral mRNA within single infected cells.
In particular, transcription of influenza virus mRNAs is performed by the viral ribonucleoprotein (RNP) complex, which consists of the three proteins that encode the tripartite polymerase (PB2, PB1, and PA) as well as nucleoprotein (NP)~\citep{huang1990determination}.
Each viral gene segment is associated with one RNP in incoming infecting virions, but secondary transcription by newly synthesized RNPs requires the presence of the viral genes encoding each of the four RNP proteins~\citep{Vreede:2004ip,eisfeld2015centre}.
This secondary transcription is a major source of viral mRNAs, as evidenced by the fact that blocking synthesis of the RNP proteins reduces the amount of viral mRNA by several orders of magnitude in bulk cells (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:cyclohexamide}).

\begin{figure}[t!]
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
\caption{\label{fig:fluburdenbyflugene}
The absence of viral genes explains some of the variability in the amount of viral mRNA per cell.
{\bf (A)} 
Fraction of mRNA in each infected cell derived from virus as a function of the normalized expression of each viral gene in that cell, taken over all time points.
Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
{\bf (B)}
Box plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each gene.
A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
Similar results are obtained if we examine only the 10-hour timepoint (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:10hrfluburdenbyflugene}).
Top and bottom of box bound the first and third quartiles; the line in the box is at the median.
Whiskers extend to the highest or lowest point within1.5$\times$ the interquartile range, and outliers are shown as circles.
{\bf (C)}
The fraction of that express each of the four other genes among cells that express all RNP genes, as well as the fraction that express \emph{all} four of the other genes.
The fraction that express all four genes is well predicted by simply multiplying the frequencies of cells that express each gene individually, indicating that gene absence is approximately independent across these genes.
}
\figsupp[Secondary transcription from newly synthesized RNPs is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
A549 cells were infected at an MOI of 0.2 as calculated on MDCK-SIAT1 cells in either the presence or absence of the protein-translation inhibitor cyclohexamide, and viral mRNA was quantified at 8 hours post-infection by qPCR.
The cyclohexamide prevents translation of new PB2, PB1, PA, and NP protein, and so prevents the formation of the new RNPs needed for secondary transcription.
The bars show the relative amount of HA and PB2 mRNA in the absence versus the presence of cyclohexamide. Error $\pm$ S.D. n=3. 
}{\includegraphics[width=0.3\linewidth]{figures/Primary_transcription_measurements/Cyclohexamide.pdf}}
\figsupp[Like panel (A), but shows samples individually.\label{figsupp:fluburdensamplebreakdown}]{
The fraction of mRNA in each infected cell derived as a function of the normalized expression of each viral gene, shown for the 10-hour and 8-hour samples individually (the other samples had too few infected cells for this analysis to be useful).
Points are colored as in Figure~\ref{fig:fluburdenbyflugene}.
}{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_by_sample.pdf}}
\figsupp[Like panel (B) but for the 10-hr sample only.\label{figsupp:10hrfluburdenbyflugene}]{
The absence of viral RNP genes but \emph{not} non-RNP genes remains significantly associated with reduced viral burden when we examine only the 10-hr sample, which is the single time point with the most data points.
The difference for NP is no longer statistically significant due to low counts of infected cells lacking NP, but the trend remains.
We do not show statistical analyses for other samples, as the number of infected cells is too low for meaningful statistical analyses.
}{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
\figdata{The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
\end{figure}

We examined the total amount of viral mRNA versus the expression of each viral gene (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}). 
Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription.
However, we observe cells that lack each of the other non-RNP genes but still derive $\approx$40\% of their mRNAs from virus, suggesting that none of the other genes are important for high levels of viral transcription.
These results are statistically supported by Figure~\ref{fig:fluburdenbyflugene}B, which shows that absence of any RNP gene but \emph{not} any other viral gene is associated with reduced amounts of viral mRNA. 
However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express(Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}). 
The rest of this variability must be due to additional mechanisms such as differences in host-cell state, viral mutations not detectable by our 3'-end sequencing approach, or inherent stochasticity.

@@

In contrast to the extreme variability in the total viral mRNA per cell, the fraction of this mRNA derived from each gene is much more consistent across cells (Figure~\ref{fig:fluexpr}A).
Total viral mRNA varies by orders of magnitude, but the fraction from any given viral gene is fairly tightly clustered around the median value for all cells (Figure~\ref{fig:fluexpr}B).
The relative levels of each viral mRNA in our cells are similar to prior bulk measurements made by Northern blots~\citep{Hatada:1989vz}, which also found an expression hierarchy of M $>$ NS $\gg$ NP $>$ NA $>$ HA $\gg$ PB2 $\sim$ PB1 $\sim$ PA.
The cell-to-cell consistency in the relative expression of different viral genes is even tighter if we limit the analysis only to cells that express all eight viral genes (Figure~\ref{fig:fluexpr}C,D).
Therefore, with the exception of complete gene absence, the factors that drive the dramatic cell-to-cell variability in the amount of viral mRNA have roughly similar effects on all viral genes in a given cell.
This finding is consistent with prior work showing positive correlations among the abundance of several viral genes in individual cells~\citep{Heldt:2015iz}.

@@

\subsection{Activation of the interferon response is rare in single infected cells.}
Because our sequencing captured all polyadenylated transcripts, we can examine whether there are prominent changes in the host-cell transcriptome in sub-populations of infected cells.
Influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes~\citep{Killip:2015dw,Iwasaki:2014dw,Crotta:2013ef}.
%In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection~\citep{Geiss:2002a}.
However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,Shalek:2013ej,Shalek:2014ey,PerezCidoncha:2014jr,bhushal2017cell}.
We therefore hypothesized that we might see two sub-populations of infected cells: one in which the interferon response inhibited viral transcription, and another in which the virus was able to express high levels of its mRNA by evading or blocking this response.

@@

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/p_cellular_heatmap.pdf}}
\caption{\label{fig:cellulargenes}
Cellular genes that co-vary in expression along with the the amount of influenza virus mRNA in individual cells expressing all eight viral genes.
Shown are all genes that are differentially expressed at a false discovery rate of 0.1. 
Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA. 
\jdbcomment{scale shows?} }
\figdata{The full results of the differential expression test is in \texttt{p\_sig\_cellular\_genes.csv}.}
\figdata{The results of a gene-set analysis are in \texttt{p\_sig\_cellular\_genes.csv}.}
\figsupp[Many genes that co-vary with the amount of viral mRNA are involved in the response to oxidative stress. \label{figsupp:ROStable}]{
Table delineating genes in (Figure~\ref{fig:cellulargenes}) that are associated with the response to oxidative stress ~\citep{Duong:2017ju,Jung:2017jz,Lee:2017ba,Peuchant:2017hj,MacLeod:2016kt,Jiang:2016cf,Gorrini:2013jv,Miura:2013fr,Kim:2009gi,Banning:2005cz,Murray:2003ie,Doyle:1999ts}.
}{\includegraphics[width=0.7\linewidth]{figures/ROStable.pdf}}
\end{figure}

Many of the genes with increased expression in cells with more viral mRNA are known or suspected to be regulated by the Nrf2 master regulator in response to oxidative stress.
These genes produce proteins that are involved in detoxification of reactive oxygen species or resultant products, the management of misfolded proteins, the electron transport chain, or a general stress response (Figure~\ref{fig:cellulargenes}-Figure~supplement~\ref{figsupp:ROStable}). 
We additionally see reduced expression of the nitric oxide synthase interacting protein (NOSIP). 
Transient oxidative stress is known to occur during viral infection, and may act in a proviral fashion via MAPK activation driving vRNP export~\citep{Amatore:2014cs}.
The antioxidant response is thought to be largely antiviral, potentially through inhibition of MAPK activity \citep{Lin:2016ec,Sgarbanti:2014ht}.
Our data do not reveal whether the expression of genes involved in the response to oxidative stress are a cause or a symptom of higher levels of viral mRNA, and further investigation of this topic is an interesting area for future work.
Overall, the expression of some of the host genes in Figure~\ref{fig:cellulargenes}, along with inherent stochasticity and specific mutations to viral genes in individual cells, probably explain most of the variation in viral mRNA levels observed across the infected cells in our dataset.

\section{Discussion}
We have quantified the total transcriptome composition of single cells infected with influenza virus.
While we observe a general increase in the amount of viral mRNA over time as expected from bulk measurements~\citep{Hatada:1989vz,Shapiro:1987ur}, there is wide variation in viral gene expression among individual infected cells.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Extreme heterogeneity of influenza virus transcription in single cells}

@@

\section{Introduction}
Viruses can cause massive and rapid changes in a cell's transcriptome as they churn out viral mRNAs and hijack cellular machinery.
For instance, cells infected with influenza virus at high multiplicity of infection (MOI) express an \emph{average} of 50,000 to 100,000 viral mRNAs per cell, corresponding to 5 to 25\% of all cellular mRNA~\citep{hatada1989control}.
In addition, infection can trigger innate-immune sensors that induce the expression of cellular anti-viral genes~\citep{Killip:2015dw,Iwasaki:2014dw,Crotta:2013ef}.
This anti-viral response is another prominent transcriptional signature of high-MOI influenza virus infection in bulk cells~\citep{Geiss:2002a}.

However, initiation of an actual influenza infection typically involves just a few virions infecting a few cells~\citep{varble2014influenza,mccrone2017evolutionary}. 
The dynamics of viral infection in these individual cells may not mirror bulk measurements made on many cells infected at high MOI.
Over 70 years ago, Max Delbruck showed that there was a $\sim$100-fold range in the number of progeny virions produced per cell by clonal bacteria infected with clonal bacteriophage~\citep{Delbruck:1945wo}.
Subsequent work has shown similar heterogeneity during infection with other viruses~\citep{Zhu:2009ea,Schulte:2014br,combe2015single,Akpinar:2016gt}, including influenza virus~\citep{Heldt:2015iz}.

In the case of influenza virus infection, targeted single-cell measurements of specific proteins or RNAs have shed light on some factors that contribute to cell-to-cell heterogeneity. 
The influenza virus genome consists of eight negative-sense RNA segments, and many infected cells fail to express one more of these RNAs~\citep{Heldt:2015iz,Dou:2017cp} or their encoded proteins~\citep{Brooke:2013kb}.
In addition, the activation of innate-immune responses is inherently stochastic~\citep{shalek2013single,shalek2014single,bhushal2017cell}, and only some influenza-infected cells express anti-viral interferon genes~\citep{PerezCidoncha:2014jr,Killip:2017ef}.
However, the extent of cell-to-cell variation in these and other host and viral factors remains unclear, as does the association among them in individual infected cells.

@@

Infected cells will express viral as well as cellular mRNAs -- however the cell barcodes and UMIs cannot distinguish whether a cell was initially infected by one or multiple viral particles.
We therefore engineered an influenza virus (strain A/WSN/1933) that additionally carried \emph{viral barcodes} consisting of synonymous mutations near the 3' end of each transcript (Figure~\ref{fig:workflow}A).
Critically, these synonymous mutations did not greatly impact viral growth kinetics (Figure~\ref{fig:workflow}B).
We infected A549 human lung carcinoma cells with an equal mix of the wild-type and synonymously barcoded viruses.
Cells infected by a single virion will exclusively express mRNAs from either wild-type or synonymously barcoded virus, whereas cells that are co-infected with multiple virions will often express mRNAs from both the wild-type and synonymously barcoded viruses (Figure~\ref{fig:workflow}C).
\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{figures/Workflow/workflow.pdf}}
\caption{\label{fig:workflow} Experimental design.
{\bf (A)}  
We engineered a virus that carried two synonymous mutations near the 3' end of each mRNA.
At top are the mutations for PB2.
At bottom are locations of the synonymous mutations relative to the typical distribution of read depth for our 3'-end sequencing.
{\bf (B)} 
The wild-type and synonymously barcoded viruses transcribe their genes with similar kinetics. 
The abundance of the viral hemagglutinin (HA) transcript relative to the cellular housekeeping gene L32 was assessed by qPCR in A549 cells infected at an MOI of 0.5 (as determined on MDCK-SIAT1 cells).
Error bars $\pm$ S.D., n=3.
{\bf (C)}  
For the single-cell mRNA sequencing, A549 cells were infected with an equal mixture of mutant and wild-type virus. 
Immediately prior to collection, cells were physically separated into droplets and cDNA libraries were generated containing the indicated barcodes. 
The libraries were deep sequenced, and the data processed to create a matrix that gives the number of molecules of each transcript observed in each cell.
Infected cells were further annotated by whether their viral mRNAs derived from wild-type virus, synonymously barcoded virus, or both.
}
\figdata{Sequences of wild-type and barcoded viruses are in \texttt{viralsequences.fasta}.\label{figdata:viralsequences}}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/validating_populations_D02.pdf}}
\caption{\label{fig:viruspopulations} The viral stocks in our experiments are relatively pure of defective particles. 
{\bf (A)}
Our viral stocks have a higher ratio of infectious particles to HA virion RNA compared to a high-defective stock propagated at high MOI.
HA viral RNA was quantified by qPCR on virions. 
Error bars $\pm$ S.D., n=6 (qPCR replicates). 
{\bf (B)} 
Our viral stocks have a higher ratio of infectious particles to particles capable of expressing HA protein.
A549 cells were infected at an MOI of 0.1, and the percentage of cells expressing HA protein at 9 hours post-infection was quantified by antibody staining and flow cytometry.
{\bf (C)} 
Our viral stocks are less immunostimulatory than virus propagated at high MOI. 
Measurements of \textit{IFNB1} transcript by qPCR normalized to the housekeeping gene L32 in A549 cells at 10 hours post infection at an MOI of 0.5.
Error bars $\pm$ S.D., n=3.
Note that MOIs were calculated by TCID50 on MDCK-SIAT1 cells, whereas the experiments in this figure involved infection of A549 cells.
}
\figsupp[Full flow cytometry data for panel B . \label{figsupp:HAstain}]{
Full flow cytometry date for  Figure~\ref{fig:viruspopulations}B. 
A549 cells were infected at an MOI of 0.1 as calculated by TCID50 on MDCK-SIAT1 cells. 
{\bf (A)} Uninfected gating control.
{\bf (B)} Cells infected with the wild-type virus stock used in our experiments.
{\bf (C)} Cells infected with synonymously barcoded virus stock used in our experiments.
{\bf (D)} Cells infected with a stock of wild-type virus propagated at a high MOI, and therefore enriched in defective particles.
}{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/HA_stain_supplement.png}}\end{figure}

@@

\subsection{Absence of viral genes partially explains cell-to-cell variability in viral load.}
The influenza genome is segmented, and cells can fail to express a viral mRNA if the encoding gene segment is not packaged in the infecting virion or fails to initiate transcription after infection.
Indeed, several groups have reported that the majority of infected cells fail to express at least one viral gene~\citep{Brooke:2013kb, Heldt:2015iz,Dou:2017cp}. 
We wondered if the absence of specific viral genes might be associated with reduced amounts of viral mRNA within single infected cells.
In particular, transcription of influenza virus mRNAs is performed by the viral ribonucleoprotein (RNP) complex, which consists of the three proteins that encode the tripartite polymerase (PB2, PB1, and PA) as well as nucleoprotein (NP)~\citep{huang1990determination}.
Each viral gene segment is associated with one RNP in incoming infecting virions, but secondary transcription by newly synthesized RNPs requires the presence of the viral genes encoding each of the four RNP proteins~\citep{Vreede:2004ip,eisfeld2015centre}.
This secondary transcription is a major source of viral mRNAs, as evidenced by the fact that blocking synthesis of the RNP proteins reduces the amount of viral mRNA by several orders of magnitude in bulk cells (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:cyclohexamide}).

\begin{figure}[t!]
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
\caption{\label{fig:fluburdenbyflugene}
The absence of viral genes explains some of the variability in the amount of viral mRNA per cell.
{\bf (A)} 
Fraction of mRNA in each infected cell derived from virus as a function of the normalized expression of each viral gene in that cell, taken over all time points.
Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
{\bf (B)}
Box plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each gene.
A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
Similar results are obtained if we examine only the 10-hour timepoint (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:10hrfluburdenbyflugene}).
Top and bottom of box bound the first and third quartiles; the line in the box is at the median.
Whiskers extend to the highest or lowest point within1.5$\times$ the interquartile range, and outliers are shown as circles.
{\bf (C)}
The fraction of that express each of the four other genes among cells that express all RNP genes, as well as the fraction that express \emph{all} four of the other genes.
The fraction that express all four genes is well predicted by simply multiplying the frequencies of cells that express each gene individually, indicating that gene absence is approximately independent across these genes.
}
\figsupp[Secondary transcription is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
A549 cells were infected at an MOI of 0.2 as calculated on MDCK-SIAT1 cells in either the presence or absence of the protein-translation inhibitor cyclohexamide, and viral mRNA was quantified at 8 hours post-infection by qPCR.
The cyclohexamide prevents translation of new PB2, PB1, PA, and NP protein, and so prevents the formation of the new RNPs needed for secondary transcription.
The bars show the relative amount of HA and PB2 mRNA in the absence versus the presence of cyclohexamide. Error $\pm$ S.D. n=3. 
}{\includegraphics[width=0.3\linewidth]{figures/Primary_transcription_measurements/Cyclohexamide.pdf}}
\figsupp[Like panel (A), but shows samples individually.\label{figsupp:fluburdensamplebreakdown}]{
The fraction of mRNA in each infected cell derived as a function of the normalized expression of each viral gene, shown for the 10-hour and 8-hour samples individually (the other samples had too few infected cells for this analysis to be useful).
Points are colored as in Figure~\ref{fig:fluburdenbyflugene}.
}{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_by_sample.pdf}}
\figsupp[Like panel (B) but for the 10-hr sample only.\label{figsupp:10hrfluburdenbyflugene}]{
The absence of viral RNP genes but \emph{not} non-RNP genes remains significantly associated with reduced viral burden when we examine only the 10-hr sample, which is the single time point with the most data points.
The difference for NP is no longer statistically significant due to low counts of infected cells lacking NP, but the trend remains.
We do not show statistical analyses for other samples, as the number of infected cells is too low for meaningful statistical analyses.
}{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
\figdata{The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
\end{figure}

We examined the total amount of viral mRNA versus the expression of each viral gene (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}). 
Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription.
However, we observe cells that lack each of the other non-RNP genes but still derive $\approx$40\% of their mRNAs from virus, suggesting that none of the other genes are important for high levels of viral transcription.
These results are statistically supported by Figure~\ref{fig:fluburdenbyflugene}B, which shows that absence of any RNP gene but \emph{not} any other viral gene is associated with reduced amounts of viral mRNA. 
However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express.
Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}). 
The rest of this variability must be due to additional mechanisms such as differences in host-cell state, viral mutations not detectable by our 3'-end sequencing approach, or inherent stochasticity.

@@

In contrast to the extreme variability in the total viral mRNA per cell, the fraction of this mRNA derived from each gene is much more consistent across cells (Figure~\ref{fig:fluexpr}A).
Total viral mRNA varies by orders of magnitude, but the fraction from any given viral gene is fairly tightly clustered around the median value for all cells (Figure~\ref{fig:fluexpr}B).
The relative levels of each viral mRNA in our cells are similar to prior bulk measurements made by Northern blots~\citep{hatada1989control}, which also found an expression hierarchy of M $>$ NS $\gg$ NP $>$ NA $>$ HA $\gg$ PB2 $\sim$ PB1 $\sim$ PA.
The cell-to-cell consistency in the relative expression of different viral genes is even tighter if we limit the analysis only to cells that express all eight viral genes (Figure~\ref{fig:fluexpr}C,D).
Therefore, with the exception of complete gene absence, the factors that drive the dramatic cell-to-cell variability in the amount of viral mRNA have roughly similar effects on all viral genes in a given cell.
This finding is consistent with prior work showing positive correlations among the abundance of several viral genes in individual cells~\citep{Heldt:2015iz}.

@@

\subsection{Activation of the interferon response is rare in single infected cells.}
Because our sequencing captured all polyadenylated transcripts, we can examine whether there are prominent changes in the host-cell transcriptome in sub-populations of infected cells.
Influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes~\citep{Killip:2015dw,Iwasaki:2014dw,Crotta:2013ef}.
%In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection~\citep{Geiss:2002a}.
However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,shalek2013single,shalek2014single,PerezCidoncha:2014jr,bhushal2017cell}.
We therefore hypothesized that we might see two sub-populations of infected cells: one in which the interferon response inhibited viral transcription, and another in which the virus was able to express high levels of its mRNA by evading or blocking this response.

@@

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/p_cellular_heatmap.pdf}}
\caption{\label{fig:cellulargenes}
Cellular genes that co-vary in expression along with the the amount of influenza virus mRNA in individual cells expressing all eight viral genes.
Shown are all genes that are differentially expressed at a false discovery rate of 0.1. 
Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA. 
\jdbcomment{scale shows?} }
\figdata{The full results of the differential expression test is in \texttt{p\_sig\_cellular\_genes.csv}.}
\figdata{The results of a gene-set analysis are in \texttt{p\_sig\_cellular\_genes.csv}.}
\figsupp[Many genes that co-vary with viral load are involved in the oxidative stress response. \label{figsupp:ROStable}]{
Table delineating genes in Figure~\ref{fig:cellulargenes} that are associated with the response to oxidative stress~\citep{Duong:2017ju,Jung:2017jz,Lee:2017ba,Peuchant:2017hj,MacLeod:2016kt,Jiang:2016cf,Gorrini:2013jv,Miura:2013fr,Kim:2009gi,Banning:2005cz,Murray:2003ie,Doyle:1999ts}.
}{\includegraphics[width=0.7\linewidth]{figures/ROStable.pdf}}
\end{figure}

Many of the genes with increased expression in cells with more viral mRNA are known or suspected to be regulated by the Nrf2 master regulator in response to oxidative stress.
These genes produce proteins that are involved in detoxification of reactive oxygen species or resultant products, the management of misfolded proteins, the electron transport chain, or a general stress response (Figure~\ref{fig:cellulargenes}-Figure~supplement~\ref{figsupp:ROStable}). 
We additionally see reduced expression of the nitric oxide synthase interacting protein (NOSIP). 
Transient oxidative stress is known to occur during viral infection, and may act in a proviral fashion via MAPK activation driving vRNP export~\citep{Amatore:2014cs}.
The antioxidant response is thought to be largely antiviral, potentially through inhibition of MAPK activity \citep{Lin:2016ec,Sgarbanti:2014ht}.
Our data do not reveal whether the expression of genes involved in the response to oxidative stress are a cause or a symptom of higher levels of viral mRNA, and further investigation of this topic is an interesting area for future work.
%Overall, the expression of some of the host genes in Figure~\ref{fig:cellulargenes}, along with inherent stochasticity and specific mutations to viral genes in individual cells, probably explain most of the variation in viral mRNA levels observed across the infected cells in our dataset.

\section{Discussion}
We have quantified the total transcriptome composition of single cells infected with influenza virus.
While we observe a general increase in the amount of viral mRNA over time as expected from bulk measurements~\citep{hatada1989control,Shapiro:1987ur}, there is wide variation in viral gene expression among individual infected cells.

The most obvious form of heterogeneity is the complete failure of some infected cells to express one or more viral genes, which we estimate occurs in about half the infected cells in our experiments.
The absence of some viral genes in some infected cells has been noted previously~\citep{Brooke:2013kb,Heldt:2015iz,Dou:2017cp}, and our work provides a holistic view by quantifying the the total viral transcriptional load as a function of the level of each mRNA.
We find that absence of any of the four genes that encode the viral RNP leads to a dramatic drop in the total amount of viral mRNA, consistent with prior bulk studies~\citep{Vreede:2004ip,eisfeld2015centre}.
Interestingly, the \emph{reason} some cells fail to express some viral genes remains unclear.
The prototypical influenza virion packages one copy of each of the eight gene segments~\citep{noda2006architecture,hutchinson2010genome}, but some virions surely package fewer~\citep{brooke2014influenza}.
However, it is also possible that much of the viral gene absence is due to stochastic loss of viral RNPs after infection but prior to the initiation of viral transcription in the nucleus.
%Further work is needed to distinguish between these possibilities.
